---
title: "Analyzing the Distribution of Unemployment Rate by Various Factors"
author: "Ruihan Zhang, Nora Zhou, Mandy Zhou, Xipu Wang"
date: "2024-10-10"
output: html_document
---

## Introduction

### Background
The COVID-19 pandemic caused significant disruptions to the global economy, leading to widespread job losses. While there has been some recovery in the U.S. job market, unemployment remains a critical issue. The economic downturn left many individuals jobless, and the number of available job openings remains limited due to the slow economic recovery. These employment issues significantly impact the workforce, particularly vulnerable groups such as low-income workers, young graduates, and those in specific sectors like hospitality and retail.

To fully understand the state of the labor market post-COVID-19, it is crucial to analyze the unemployment rate and the factors influencing employment in 2021.

### Purpose
The primary objective of this project is to analyze the 2021 unemployment rate across the United States, focusing on the factors that influence employment outcomes. Using data from the IPUMS USA, which harmonizes U.S. census microdata, we aim to explore key variables such as age, education, gender, industry, geographic location, and other work-related factors that contribute to unemployment status. This analysis seeks to understand how these factors shaped the employment landscape during a period of economic uncertainty and slow recovery.

### Objectives
1. Understand the general distribution of key demographic variables such as age, sex, and gender in the workforce.
2. Analyze the geographical distribution of unemployment across different U.S. states.
3. Investigate the relationship between the unemployment rate and various work-related factors, including education levels and working conditions.

## Dataset Description

### Source
The data for this analysis is sourced from IPUMS USA, which collects, preserves, and harmonizes U.S. census microdata, providing easy access with enhanced documentation. The dataset includes decennial censuses from 1790 to 2010 and American Community Surveys (ACS) from 2000 to the present, offering detailed information on various demographic and employment-related variables.

### Scope
For this analysis, we focus on data from the year 2021 to conduct a post-COVID-19 examination of the labor market. Relevant variables include demographic data (age, sex, gender), employment-related factors (such as education and working conditions), disability status, salary, income, and other key factors. After extracting the original dataset, we narrowed our focus to specific objectives and selected the most important variables for an in-depth analysis.

### Important Variables

| Variable Name | Interpretation |
|---------------|----------------|
| `STATEICP`    | Identifies the state in which the housing unit was located using the coding scheme developed by the Inter-University Consortium for Political and Social Research (ICPSR). |
| `Age`         | Reports the person's age in years as of the last birthday. |
| `sex`         | Reports whether the person was male or female. |
| `Race`        | Indicates the total number of major race groups reported for an individual. |
| `EDUC`        | Indicates respondents' educational attainment as measured by the highest year of school or degree completed. |
| `EMPSTAT`     | Indicates whether the respondent was part of the labor force—working or seeking work—and, if so, whether the person was currently unemployed. |
| `OCC2010`     | Offers a consistent long-term classification of occupations. |
| `LABFORCE`    | A dichotomous variable indicating whether a person participated in the labor force. |
| `DIFFSENS`    | Indicates whether the respondent has a long-lasting condition of blindness, deafness, or a severe vision or hearing impairment. |
| `VETSTAT`     | Indicates whether individuals served in the U.S. military forces during times of war or peace. |
| `UHRSWORK`    | Reports the number of hours per week that the respondent usually worked if they worked during the previous year. |

## Analysis



```{r}
library(tidyverse)
library(maps)
library(ggplot2)
library(readr)
library(viridis)
library(dplyr)

```

```{r}
cleaned_data <- read_csv("/Users/zhangruihan/Desktop/QTM 302W/cleaned_data.csv")

```

# Education

```{r}
# Summarize unemployment by education level
education_unemployment <- cleaned_data %>%
  group_by(EDUC, EMPSTAT) %>%
  summarise(count = n()) %>%
  spread(key = EMPSTAT, value = count, fill = 0) %>%
  rename(
    employed = `1`,
    unemployed = `2`
  )

education_unemployment <- education_unemployment %>%
  mutate(total = employed + unemployed,
         unemployment_rate = unemployed / total)

print(education_unemployment)

```

```{r}
ggplot(education_unemployment, aes(x = as.factor(EDUC), y = unemployment_rate)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(
    title = "Unemployment Rate by Education Level",
    x = "Education Level",
    y = "Unemployment Rate",
    caption = "Source: Cleaned Data"
  ) +
  theme_minimal()

```

```{r}
# Create a data frame for education values and labels
education_table <- data.frame(
  Value = c("00", "01", "02", "03", "04", "05", "06", "07", "08", "09", "10", "11", "99"),
  Label = c(
    "N/A or no schooling",
    "Nursery school to grade 4",
    "Grade 5, 6, 7, or 8",
    "Grade 9",
    "Grade 10",
    "Grade 11",
    "Grade 12",
    "1 year of college",
    "2 years of college",
    "3 years of college",
    "4 years of college",
    "5+ years of college",
    "Missing"
  )
)

print(education_table)
```

Peak Unemployment Around High School Level:
The unemployment rate appears to peak around education levels 4 and 5, which likely correspond to Grade 10 and Grade 11. This could indicate that individuals who do not complete high school face greater barriers to employment compared to those with higher levels of education.
Declining Unemployment with Higher Education:
As the education level increases beyond high school (levels 7 to 11, representing college education), the unemployment rate generally decreases. This trend suggests that higher education provides better employment prospects, potentially due to increased skills and qualifications.
The lowest unemployment rate is observed for those with 5+ years of college (level 11), indicating that advanced education might be associated with higher job security.

# Race

```{r}
# Summarize unemployment by education level
race_unemployment <- cleaned_data %>%
  group_by(RACE, EMPSTAT) %>%
  summarise(count = n()) %>%
  spread(key = EMPSTAT, value = count, fill = 0) %>%
  rename(
    employed = `1`,
    unemployed = `2`
  )

race_unemployment <- race_unemployment %>%
  mutate(total = employed + unemployed,
         unemployment_rate = unemployed / total)

print(race_unemployment)

```

```{r}
ggplot(race_unemployment, aes(x = as.factor(RACE), y = unemployment_rate)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(
    title = "Unemployment Rate by Race",
    x = "Race",
    y = "Unemployment Rate",
    caption = "Source: Cleaned Data"
  ) +
  theme_minimal()
```

```{r}
# Create a data frame for race values and labels
race_table <- data.frame(
  Value = c("1", "2", "3", "4", "5", "6", "7", "8", "9"),
  Label = c(
    "White",
    "Black/African American",
    "American Indian or Alaska Native",
    "Chinese",
    "Japanese",
    "Other Asian or Pacific Islander",
    "Other race, nec",
    "Two major races",
    "Three or more major races"
  )
)

# Display the race table
print(race_table)

```
Higher Unemployment Rates for Certain Racial Groups:
The bars for races with codes 2 (Black/African American) and 3 (American Indian or Alaska Native) show notably higher unemployment rates compared to other racial groups, with rates over 10%.
This suggests that individuals in these racial categories face greater challenges in securing employment, potentially reflecting systemic disparities in the labor market.
Relatively Lower Unemployment Rates for Other Groups:
The unemployment rate for individuals identified as race 1 (White) is among the lower rates, around 5%. This indicates that this group might have better access to employment opportunities compared to other racial categories in the dataset.
Races 4 (Chinese) and 5 (Japanese) also show relatively lower unemployment rates, suggesting that these groups may have different economic opportunities or access to resources that support employment.
Mixed Results for Multiracial Categories:
The unemployment rates for races 8 (Two major races) and 9 (Three or more major races) are higher compared to some of the single-race categories. This could indicate that individuals identifying with multiple racial groups face unique challenges in the job market.


# Sex

```{r}
sex_data <- cleaned_data %>%
  mutate(
    SEX = factor(SEX, levels = c(1, 2, 9), labels = c("Male", "Female", "Missing/Blank")),
    EMPSTAT = factor(EMPSTAT, levels = c(1, 2), labels = c("Employed", "Unemployed"))
  )
```

```{r}
sex_unemployment <- sex_data %>%
  filter(SEX %in% c("Male", "Female")) %>%  
  group_by(SEX, EMPSTAT) %>%
  summarise(count = n(), .groups = 'drop') %>%  
  mutate(percentage = count / sum(count) * 100)  

print("Sex vs. Employment Status Percentage Data:")
print(sex_unemployment)

ggplot(sex_unemployment, aes(x = EMPSTAT, y = percentage, fill = SEX)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Sex vs. Employment Status (Percentage)", x = "Employment Status", y = "Percentage (%)") +
  theme_minimal() +
  scale_fill_manual(values = c("Male" = "lightblue", "Female" = "pink")) 


```

The visualization shows that a slightly higher percentage of males are employed compared to females. Conversely, the percentage of females in the unemployed category is slightly higher than that of males. This indicates a potential gender disparity in employment status, where females may face slightly higher unemployment rates compared to males.

# Distribution across states

```{r}
state_codes <- c(
  `1` = 'Connecticut',
  `2` = 'Maine',
  `3` = 'Massachusetts',
  `4` = 'New Hampshire',
  `5` = 'Rhode Island',
  `6` = 'Vermont',
  `11` = 'Delaware',
  `12` = 'New Jersey',
  `13` = 'New York',
  `14` = 'Pennsylvania',
  `21` = 'Illinois',
  `22` = 'Indiana',
  `23` = 'Michigan',
  `24` = 'Ohio',
  `25` = 'Wisconsin',
  `31` = 'Iowa',
  `32` = 'Kansas',
  `33` = 'Minnesota',
  `34` = 'Missouri',
  `35` = 'Nebraska',
  `36` = 'North Dakota',
  `37` = 'South Dakota',
  `40` = 'Virginia',
  `41` = 'Alabama',
  `42` = 'Arkansas',
  `43` = 'Florida',
  `44` = 'Georgia',
  `45` = 'Louisiana',
  `46` = 'Mississippi',
  `47` = 'North Carolina',
  `48` = 'South Carolina',
  `49` = 'Texas',
  `51` = 'Kentucky',
  `52` = 'Maryland',
  `53` = 'Oklahoma',
  `54` = 'Tennessee',
  `56` = 'West Virginia',
  `61` = 'Arizona',
  `62` = 'Colorado',
  `63` = 'Idaho',
  `64` = 'Montana',
  `65` = 'Nevada',
  `66` = 'New Mexico',
  `67` = 'Utah',
  `68` = 'Wyoming',
  `71` = 'California',
  `72` = 'Oregon',
  `73` = 'Washington',
  `81` = 'Alaska',
  `82` = 'Hawaii',
  `83` = 'Puerto Rico',
  `96` = 'State groupings'
)
```


```{r}
cleaned_data$State <- state_codes[as.character(cleaned_data$STATEICP)]
cleaned_data$Unemployed <- ifelse(cleaned_data$EMPSTAT == 2, 1, 0)
state_unemployment <- cleaned_data %>%
  group_by(State) %>%
  summarize(
    Total_Population = n(),
    Total_Unemployed = sum(Unemployed, na.rm = TRUE)
  ) %>%
  mutate(
    Unemployment_Rate = (Total_Unemployed / Total_Population) * 100
  )
```


```{r}
states_map <- map_data("state")
```


```{r}
state_unemployment <- state_unemployment %>%
  mutate(region = tolower(State))
map_data <- left_join(states_map, state_unemployment, by = "region")
ggplot(map_data, aes(x = long, y = lat, group = group, fill = Unemployment_Rate)) +
  geom_polygon(color = "white") +
  coord_fixed(1.3) +
  scale_fill_viridis(
    option = "magma",
    na.value = "grey80",
    name = "Unemployment Rate (%)"
  ) +
  theme_minimal() +
  labs(
    title = "Unemployment Rate by State",
    caption = "Source: Cleaned Datas"
  ) +
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    legend.position = "right"
  )
```

From the map that is created above, we can see that for mid-north states like Montana, North Dakota, and Nebraska, the unemployment rate is significantly low (below 4%). But for states at the south-west, like Florida, Georgia, and Louisiana, the unemplotment rate is quite high (around 5%). The west-coast states have a significantly high unemployment rate, with California at 8%, Las Vegas over 8%.

# Age

```{r}
cleaned_data$AgeGroup <- cut(
  cleaned_data$AGE,
  breaks = seq(0, 100, by = 10),
  right = FALSE,
  labels = paste(seq(0, 90, by = 10), seq(9, 99, by = 10), sep = "-")
)

```

```{r}
age_unemployment <- cleaned_data %>%
  filter(!is.na(AgeGroup)) %>%
  group_by(AgeGroup) %>%
  summarize(
    Total_Population = n(),
    Total_Unemployed = sum(Unemployed, na.rm = TRUE)
  ) %>%
  mutate(
    Unemployment_Rate = (Total_Unemployed / Total_Population) * 100
  )

```

```{r}
ggplot(data = age_unemployment, aes(x = AgeGroup, y = Unemployment_Rate)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme_minimal() +
  labs(
    title = "Unemployment Rate by Age Group",
    x = "Age Group",
    y = "Unemployment Rate (%)"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

Higher Unemployment Rates Among Younger Age Groups:
The highest unemployment rate is observed in the 10-19 age group, which exceeds 12%. This is typical, as many individuals in this age range are new to the workforce and may lack experience or face competition for entry-level positions.
The unemployment rate decreases for the 20-29 age group but remains relatively high compared to older age groups, likely due to continued career establishment and competition in the job market.
Stabilization in Middle Age Groups:
Unemployment rates for the 30-39, 40-49, and 50-59 age groups show a significant decline compared to younger groups, falling below 6%.
This trend may reflect increased job stability and career advancement as individuals gain more experience and secure more stable positions.
Slight Increase in Unemployment for Older Age Groups:
There is a noticeable increase in unemployment rates for the 70-79 age group, possibly as older workers face challenges such as age discrimination or skill mismatches in the evolving job market.
However, unemployment rates decline again for the 80-89 and 90-99 age groups, likely reflecting that many individuals in these age ranges are retired and not actively seeking employment.